---
title: "Psychology of Survey Response"
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Factor Analysis to Understand Political Emotions

We will conduct both exploratory and confirmatory factor analysis using the nationally representative 2020 [American National Election Survey](https://electionstudies.org/data-center/2020-time-series-study/) for our data. In this survey, 9 unique political emotions variables were asked of respondents, and we will use these 9 variables to understand how political emotions are structured. Currently, it is widely believed that three latent emotional factors best explain the structure of political emotions [@marcus2006measure] with three primary latent factors: 

1. Aversion to Politics: Measures include anger, outrage, and irritation 
2. Worry about Politics: Measures include fear, worry, and nervousness 
3. Enthusiasm about Politics: Measures include happiness, hope, and pride

We will use factor analysis to test whether a three factor solution really is the best way to explain these political emotions.

To start, we read in our data, in this case the 2020 American National Election study. Then we save a new data frame that includes only the nine political emotions variables we want to include in our analysis. We will use this data frame throughout the code. Once we have saved the political emotions variable as a new data frame, we recode all the negative values to _NA_ as these are non-substantive responses which should be removed (as indicated in the associated codebook).

Next, we change the variables names to something that is more informative. This will aid in interpretation of the factor analysis results and should not be skipped. Lastly, we use the `skimr` package to quickly skim the variables in our dataset. We want to see the minimum value = 1 with no negative values, since negative values should be treated as missing data.

```{r}

library(haven) #Imports stata file
library(skimr) #For data evaluation 
library(psych) #For exploratory factor analysis
library(corrplot) #To graph correlations 
library(lavaan) #For confirmatory factor analysis
library(semPlot) #For graphing CFA results 
library(GPArotation) #For factor rotation


anes <- read_dta('C:/Users/Stefani.Langehennig/OneDrive - University of Denver/Documents/research/surveys-textbook-data/anes_timeseries_2020_stata_20220210.dta')

df <- data.frame(anes$V201115, anes$V201116, anes$V201117, anes$V201118, anes$V201119, anes$V201120, anes$V201121, anes$V201122, anes$V201123  ) #Save the variables you want to include in your factor analysis into a new data frame

df[df <= -1] <- NA #Set missing values to NA to not include in analysis

new_names <- c("hope", "afraid", "outrage", "angry", "happy", "worried", "proud", "irritated", "nervous") #Give your variables new informative names 

# Update column names
colnames(df) <- new_names #Apply new names to your data frame

skim(df) #Checks the variables in your data frame; evaluate for missing data

```

#### Step 1: Evaluating Correlations Between Political Emotions

First, we need to check the correlations between your variables - here, political emotions - to see how related each of the individual items are. Below, we create a matrix with all the correlations between the individual items and graph the correlations using a heat-map for easier viewing.

```{r }
#Step 1: Evaluate correlations 
cor_matrix<-cor(df, use = "pairwise.complete.obs") #Saves correlation matrix

corrplot(cor_matrix, method = "circle") # Plot correlation matrix as a heatmap

cor_matrix
###At least 2 distinct and possibly 3 distinct factors from examining 
```

- *Results indicate 2 and probably 3 unique factors +* Positive Emotions: Happy, Pride, Hope

Results indicate at least two and likely three distinct factors in the political emotions' variables. Three factors would match the dominant belief in the literature on how political emotions are structured [@marcus2006measure]. The three positive emotions are clearly positively related to each other with smaller and negative coefficients to the other six emotions. For the negative emotions, all six items are positively and significantly related to one another indicating they might all be measuring the same concept. However, closer examination of the results indicates that being afraid is more highly correlated with being worried or nervous than it is anger, outrage, or irritation. That matches the underlying theoretical belief that those three items represent an "anxiety about politics" factor, whereas anger, outrage, and irritation represent an "aversion" factor towards politics.

Conducting a factor analysis will help us better understand how these 9 individual political emotions are related to one another. We will conduct both exploratory and confirmatory factor analysis to illustrate these methods.

### Exploratory Factor Analysis

We'll start the factor analysis with an exploratory factor analysis approach using both principal component factor analysis (pcf) and principal axis factor analysis (paf). We'll compare the results between these two approaches to evaluate differences. PCF handles non-linearity better than PAF approaches, so with many Likert style survey questions that are not normally distributed we utilize the PCF approach.

#### Step 2: Screeplot

```{r scree}
#Step 2: Evaluate Screeplot - looking for number of factors >= ~1 
scree(df) #from 'psych' package and graphs scree plot for PCF and PAF approaches 
```

::: {.callout-tip}
## Scree Plots
We use scree plots to visually decide how many factors or components to use in our factor analyses. We usually determine this number by looking at everything that precedes the elbow.
:::

The scree plot shows eigenvalues from a PCF and a PAF, non-rotated, factor analysis. The PCF shows two clear factors with a third worth looking into whereas the PAF shows one clear factor with a second factor that is close. Knowing what we know from the correlations we examined, we will start with a three factor solution with our exploratory analysis. If two factors does indeed fit the data better than three, the factor analysis will show that.

#### Step 3: Estimate the Exploratory Factor Analysis

Now, we will estimate a series of factor analyses to illustrate PCF vs. PAF Rotation Types: None, Orthogonal, & Oblique

Note the slight difference in code between the `pcf` and the `paf` approaches. The only difference is using `principal` for the `pcf` approach and `fa` for the `paf`. We also can indicate which form of rotation we want to perform for each of the factor analyses. Here, we do both no rotation and use varimax for the orthogonal rotations, and oblimin for the oblique rotations. We start with a basic three factor pcf approach without rotation.

```{r, pcf estimate}

pcf_result_no <- principal(df,nfactors = 3,  rotate = "none") #PCF approach with no rotation

pcf_result_no

```

Let's evaluate the results. The first thing to review is the `SS loadings` row of results. The three values shown in that row are the eigenvalues for the 3 unique factors we specified. The first two factors both have eigenvalues \>1 while the third factor's eigenvalue is not quite at 1. We also want to evaluate the proportion of the variance that each factor explains. Factor 1 clearly explains the most (\~61%) while the third factor only adds 7% of additional explained variance.

Next, we'll review the actual factors and see which measures load on which factor. We see that the six negative emotions all seem to load on Factor 1 while the three positive emotions seem to load on Factor 2. The third factor seems to loosely be related to political anxiety and includes being afraid, worried, and nervous. While Factor 3 is not clearly unique in the unrotated factor analysis, the fact that there are reasonably strong factor loadings indicates that rotation may help to reveal a clearer pattern in the results.

::: {.callout-important}
Once you assess which measures load onto which factors, always give the factors a substantive name that is intuitive.
:::

```{r}
fa.diagram(pcf_result_no) #Graphs the relationship
```

Finally, we can also graph the factor results and see that in the unrotated results all nine emotions load most strongly on Factor 1, even though the positive emotions and negatively related to the negative emotions. This graph takes the absolute value of the factor loadings and matches the highest factor loading for that item to the appropriate latent factor.

With our knowledge of the correlation matrix and the strong factors loadings from the unrotated pcf model, we will go ahead and rotate our factor analysis results. We will use an orthogonal rotation, varimax, which removes all shared variance between the latent factors.

```{r}

pcf_result_var <- principal(df,nfactors = 3,  rotate = "varimax") #PCF approach with varimax rotation

pcf_result_var #Rotation reveals cleaner factors that are obscured 

fa.diagram(pcf_result_var) #Graphs the relationship 

```

Let's review the same three items in this analysis. Starting with the `SS loadings` we see that the values are much more evenly aligned across the factors with eigenvalues \>=1 for all three factors. Because of this, the proportion of total variance explained is also very close across the three factors. This indicates that the rotation was needed and successful in revealing a clearer pattern in the data. There are likely three factors based on these results.

Next, by examining the factor loadings for each latent factor, we see that the results follow our theoretical beliefs. Anger, outrage, and irritation loaded on Factor 1; hope, happiness, and pride loaded together on Factor 2; and anxiety, worry, and nervousness loaded on Factor 3. When we graph the results we see that clear pattern as well. Generally, the conclusion seems to be that political emotions are structured into three unique latent factors.

The following code compares the pcf approach used above alongside the paf approach, as well as changes the rotation time to oblimin for the oblique rotational approach.

```{r, factor analysis estimate}

#General pcf code "principal(data frame, nfactors=x, rotate)"
#General paf code "fa(data frame, nfactors=x, rotate)"

#####Principal Components Factor Analysis, 3 factor solution with no rotation, orthogonal (varimax) & oblique (oblimin)
pcf_result_no <- principal(df,nfactors = 3,  rotate = "none") #PCF approach with no rotation
pcf_result_no #Reports same Eigenvalues as reported in Scree Plot

pcf_result_var <- principal(df,nfactors = 3,  rotate = "varimax") #PCF approach with varimax rotation

pcf_result_var #Rotation reveals cleaner factors that are obscured 

pcf_result_obl <- principal(df,nfactors = 3,  rotate = "oblimin") #PCF approach with varimax rotation

pcf_result_obl #Rotation reveals cleaner factors that are obscured 

#####Principal Axis Factor Analysis, 3 factor solution with no rotation, orthogonal (varimax) & oblique (oblimin)
paf_result_no <- fa(df, nfactors = 3, rotate = "none") #paf model

paf_result_no #Reports same Eigenvalues as reported in Scree Plot

paf_result_var <- fa(df, nfactors = 3, rotate = "varimax") #paf model

paf_result_var

paf_result_obl <- fa(df,nfactors = 3,  rotate = "oblimin") #PCF approach with varimax rotation

paf_result_obl #Rotation reveals cleaner factors that are obscured 

```

The results show interesting patterns. Both unrotated models reveal nearly identical results, where as all four rotated results revealed a likely three factor solution. By drawing the same conclusions from the different approaches, that makes it more likely that the three factor solution is best fit for the results.

### Confirmatory Factor Analysis (CFA)

Because we have _a priori_ theory on the appropriate factor structure for these 9 political emotions, we can also use a confirmatory factor analysis to test that three factors does best fit the data.

The `lavaan` package is needed for this type of analysis. Because we are testing _a priori_ theory with this approach, we have to specify which items create which latent factor. First, we will use the three factor solution theory, which suggests we will find:

-   Factor 1 = Outrage, anger, and irritation
-   Factor 2 = Pride, happiness, and hope
-   Factor 3 = Being afraid, worried, and nervousness

```{r}

####Confirmatory Factor Analysis
# Load required packages
library(lavaan)

# Specify the CFA model; it must include the #
model <- '
   # Factor 1
   Factor1 =~ outrage + angry  + irritated
   
   # Factor 2
   Factor2 =~ proud + happy  + hope
   
   # Factor 3
   Factor3 =~ afraid  + nervous + worried
'

# Step 3: Fit the CFA model with varimax rotation
fit <- cfa(model, data = df)

# Step 4: Summarize the results
summary(fit, standardized = TRUE, fit.measures = TRUE)  #Gives you summary statistics of the CFA

semPaths(fit, "std", whatLabels = "est", edge.label.cex = 0.8) #Graphs the CFA factor loadings
```

There are several things to evaluate in the results for a confirmatory factor analysis. We are more concerned with model fit here than we are in with exploratory factor analysis since we are testing specific hypotheses. We will evaluate the following CFA model fit using the following parameters:

  - Root Mean Square Error (RMSEA) where lower values = better fitting model.
  - RMSEA \<=.06 considered good fit
  - Model RMSEA = .044 which is below the (arbitrary) cut point of .06 to indicate good fitting model
  - Standardized Root Mean Square Residual (SRMR) which is the standardized version of RMSEA.
    - SRMR\<=.1 considered good fit
  - Model SRMR = .018 which is below the (arbitrary) cut point of .1 to indicate a good fitting model
  - Comparative Fit Index reflects improvement in model fit compared to a null model.
    - Closer to 1 indicates better fitting model with \>=.9 considered good fit
  - Model CFI = .991 nearly 1 so indicates very good fitting model
  - Tucker Lewis Index is similar to CFI and reflects improvement in model fit compared to a null model.
    - Closer to 1 indicates better fitting model with \>=.9 considered good fit
  - Model TLI = .989 also higher than .9
  - AIC and BIC are useful for comparing across nested models so we can compare these values to a two factor solution to see which best fits the data
    - AIC = 179527;
    - BIC = 179674
  - With large sample sizes like we have here, the chi-square calculation will almost always be significant so does not provide valuable information here.

All of the model fit parameters for the three factor solution indicate a good fitting model. This provides initial support that our hypothesized latent emotions do, in fact, exist. However, we should change our model to evaluate if other factor solutions might fit the data better. We will use a two factor solution, combining the six negative emotions and three positive emotions into their own unique factors, and compare those results to the initial results.

First, we create a new model that combines the emotions in the manner previously stated. Then we review the results.

```{r}

model2 <- '
   # Factor 1
   Factor1 =~ outrage + angry  + irritated + afraid  + nervous + worried 
   
   # Factor 2
   Factor2 =~ proud + happy  + hope
   

'

# Step 3: Fit the CFA model with varimax rotation
fit2 <- cfa(model2, data = df)

# Step 4: Summarize the results
summary(fit2, standardized = TRUE, fit.measures = TRUE)

semPaths(fit2, "std", whatLabels = "est", edge.label.cex = 0.8) #Graphs the CFA factor loadings
```

Evaluating the same model fit parameters as before, we can immediately see a worse fitting model compared to the prior three factor solution.

  - RMSEA = .044 with three factor solution vs. .13 in two factor; closer to 0 the better 
  - SRMR = .018 in three factor vs. .038 in 2 factor; closer to 0 the better
  - CFI = .991 in three factor vs. .933 in two factor; closer to 1 indicates better fitting model
  - TLI = .989 in three factor vs. .907 in two factor; closer to 1 indicates better fitting model
  - AIC =179527 in three factor vs. 182734 in two factor
  - BIC =179674 in three factor vs. 182867 in two factor

Across all of the mode fit parameters, the three factor solution fits the data better than the two factor solution that combined all the negatively valenced emotions. These results match the results from the exploratory factor analysis as well.

### Conclusions

In this chapter, we imported survey data and conducted various types of factor analysis techniques on political emotions in the United States. The results largely follow the prevailing theoretical belief that there are three distinct emotional latent factors:

  - Aversion to Politics: Measures anger, outrage, and irritation
  - Worry about Politics: Measures fear, worry, and nervousness
  - Enthusiasm about Politics: Measures happiness, hope, and pride







