---
title: "Survey Analysis: Logitistic & Probit Regression"
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(ggplot2)
library(stargazer)
library(MASS)
library(tidyverse)
set.seed(1964)

```

## Analyzing Dichotomous Dependent Variables

We are going use the binomial distribution to simulate a dichotomous dependent variable, _y_, with a known relationship to an independent variable, _x_. We will then estimate the relationship using OLS, logistic, and probit regressions and compare the results. Along the way, we will talk about the differences in the modeling approaches. 

Let's start with simulating some data for our analysis.

```{r simulate data}
# Create X as a sequence from 1 to 11
X <- 1:11

# Define the probabilities for Y = 1 at each X value with steeper relationship in middle and flatter at edges
probabilities <- c(0.95, .925, .875, 0.8, .7, 0.5, 0.3, 0.2, 0.125, .075, 0.05)

# Simulate data with binomial distribution for Y
sample <- 1000  # Number of trials for each x 

# Create an empty data frame to store the results
simulated_data <- data.frame(X = numeric(),
                             Y = numeric(),
                             stringsAsFactors = FALSE)

# Simulate Y=1 probabilistically for each X value iteration
for (i in 1:sample) {
  # Sample Y=1 probabilistically at each X value
  simulated_Y <- sapply(probabilities, function(p) sample(c(0, 1), size = 1, prob = c(1 - p, p))) #Probabilistically draw 1 | 0 at the specified probability level 
  
  # Create a data frame for the current iteration
  iteration_data <- data.frame(x = X, y = simulated_Y)
  
  # Append the current iteration data to the main data frame
  simulated_data <- rbind(simulated_data, iteration_data)
}

head(simulated_data) #Look at first few cases 

```
### Reviewing Simulated Data

Using the `group_by` function from the `tidyverse` package, we check the probability that Y = 1 at each X level to see if the simulation worked as hoped. The results show that it did with P(Y=1∣X )roughly following the population parameters stipulated earlier.

```{r pressure, echo=FALSE}
simulated_data %>%
  group_by(x) %>%
  summarize(mean_y = mean(y))

```
### Estimating Regression Models

With our simulated data, now we'll estimate three separate regressions changing the Generalized Linear Model (GLM) type using: _Logistic, Probit,_ and _OLS_.

The two models, logistic and probit, are designed to work with dichotomous DVs and should work better with our simulated data compared to the OLS estimator (which assumes a normal distribution). We use GLMs to estimate all three models, including the OLS, to more easily compare model fit across estimators. Both the logitistic and probit models use the `family=binomial` argument in the code but have the different link functions to account for the different calculations. The OLS GLM model uses the _Gaussian_ distribution family with the standard _identity_ link function.

We estimate each of the three models than compare the results using the `stargazer` package. 

```{r modeling}
# Logistic regression
logit <- glm(y ~ x , data = simulated_data, family = binomial(link = "logit"))
summary(logit)

# Probit regression
probit <- glm(y ~ x , data = simulated_data, family = binomial(link = "probit"))
summary(probit)

# OLS regression
ols <- glm(y ~ x, data = simulated_data, family = gaussian(link = "identity"))
summary(ols)

# Compare results
stargazer(ols, logit, probit, type="text")
```
The results show that _x_ is a significant predictor of _y_ across all three model estimation strategies, which it should be given the data generating process that created these data. However, the coefficients are interpreted differently in each of the three models. For the OLS model, the coefficient of -.105 reflects a linear estimation of how _x_ is related to _y_ so that for every 1 unit increase in _x_, regardless of where on the _x_ scale you are, results in a .105, or 10.5%, decrease in the probability of _Y_. The OLS model is assuming a linear relationship and is creating the coefficient that best describes the linear relationship between _x_ and _y_.

The coefficients on the logistic and probit models do not assume linearity, meaning that the impact of _x_ on values of _y_ will vary depending on where on the _x_ scale you are. With _x_ values closer to the middle of the scale, the relationship will be stronger so that a 1 unit increase in _x_ results in the largest possible change in _y_ With _x_ values closer to either its maximum or minimum, the relationship between _x_ and _y_ will be weaker so that a 1 unit increase in _x_ results in a smaller change in _y_.



